Component response rate variation underlies the stability of complex systems

### 2020-03-02 notebook/ms.Rmd
Edit the notation a bit in the Introduction to make the results easier to read. Also, add some more analytical explanation to the start of the results

Randomly assembled complex systems can be represented as large square matrices ($\mathbf{M}$) with $S$ components (e.g., networks of species [@Allesina2012] or banks [@Haldane2011]). One element of such a matrix, $M_{ij}$, defines how component $j$ affects component $i$ in the system at a point of equilibrium [@Allesina2012]. Off-diagonal elements ($i \neq j$) therefore define interactions between components, while diagonal elements ($i = j$) define component self-regulation (e.g., carrying capacity in ecological communities). Traditionally, off-diagonal elements are assigned non-zero values with a probability $C$, which are sampled from a distribution with variance $\sigma^{2}$; diagonal elements are set to -1 [@May1972; @Allesina2012; @Allesina2015]. Local system stability is assessed using eigenanalysis, with the system being stable if the real parts of all eigenvalues ($\lambda$) of $\mathbf{M}$ are negative ($\max\left(\Re(\lambda)\right) < 0$) [@May1972; @Allesina2012]. In a large system (high $S$), eigenvalues are distributed uniformly [@Tao2010] within a circle centred at $\Re = -1$ (the mean value of diagonal elements) and $\Im = 0$, with a radius of $\sigma\sqrt{SC}$ [@May1972; @Allesina2012; @Allesina2015] (Figs 1a and 2a). Local stability of randomly assembled systems therefore becomes increasingly unlikely as $S$, $C$, and $\sigma^{2}$ increase.

May's [@May1972; @Allesina2012] stability criterion $\sigma\sqrt{SC} < 1$ assumes that the expected response rates ($\gamma$) of individual components to perturbations of the system are identical, but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of organisms, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance on stability has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma$. Here I show that, despite higher $\sigma$, realistic complex systems (such that $S$ is high but finite) are actually more likely to be stable if their individual component response rates vary. My results are robust across commonly observed network structures, including random [@May1972], small-world [@Watts1998], scale-free [@Albert2002], cascade food web [@Williams2000] networks.

@@

Equation \ref{defM} thereby serves as a null model to investigate how varying component response rates affect complex systems.

>>>>

Randomly assembled complex systems can be represented as large square matrices ($\mathbf{M}$) with $S$ components (e.g., networks of species [@Allesina2012] or banks [@Haldane2011]). One element of such a matrix, $M_{ij}$, defines how component $j$ affects component $i$ in the system at a point of equilibrium [@Allesina2012]. Off-diagonal elements ($i \neq j$) therefore define interactions between components, while diagonal elements ($i = j$) define component self-regulation ($d$, e.g., carrying capacity in ecological communities). Traditionally, off-diagonal elements are assigned non-zero values with a probability $C$, which are sampled from a distribution with variance $\sigma_{M}^{2}$; diagonal elements are set to $d = -1$ [@May1972; @Allesina2012; @Allesina2015]. Local system stability is assessed using eigenanalysis, with the system being stable if the real parts of all eigenvalues ($\lambda$) of $\mathbf{M}$ are negative ($\max\left(\Re(\lambda)\right) < 0$) [@May1972; @Allesina2012]. In a large system (high $S$), eigenvalues are distributed uniformly [@Tao2010] within a circle centred at $\Re = -1$ (the mean value of diagonal elements) and $\Im = 0$, with a radius of $\sigma_{M}\sqrt{SC$ [@May1972; @Allesina2012; @Allesina2015] (Figs 1a and 2a). Local stability of randomly assembled systems therefore becomes increasingly unlikely as $S_{M}$, $C_{M}$, and $\sigma_{M}^{2}$ increase.

May's [@May1972; @Allesina2012] stability criterion $\sigma_{M}\sqrt{SC} < 1$ assumes that the expected response rates ($\gamma$) of individual components to perturbations of the system are identical, but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of organisms, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance on stability has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ ($\sigma^{2}_{\gamma}$) might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma_{M}$. Here I show that, despite higher $\sigma_{M}$, realistic complex systems (such that $S$ is high but finite) are actually more likely to be stable if their individual component response rates vary. My results are robust across commonly observed network structures, including random [@May1972], small-world [@Watts1998], scale-free [@Albert2002], cascade food web [@Williams2000] networks.

@@

Equation \ref{defM} thereby serves as a null model to investigate how variation in component response rate ($\sigma_{\gamma}$) affects complex systems. In the absence of such variation ($\sigma_{\gamma} = 0$), $\gamma$ is set to the identity matrix (diagonal elements all equal 1) and $\textbf{M} = \textbf{A}$. For this baseline condition, $\max(\Re(\lambda))$ can be estimated from five values [@Tang2014b]. These values include (1) system size ($S$), (2) mean self-regulation of components ($d$), (3) mean interaction strength between components ($\mu$), (4) the standard deviation between component interaction strengths ($\sigma_{M}$), and (5) the correlation of interaction strengths between components, $M_{ij}$ and $M_{ji}$ ($\rho$).

When investigating the effect of varying component response rate $\sigma_{\gamma}$ on stability by defining $\textbf{M} = \gamma\textbf{A}$, $S$ remains unchanged. Further, values of $\gamma_{i}$ were sampled such that $E[d]$ and $E[\mu]$ also remained unchanged (in practice, diagonal elements of $\textbf{M}$ were standardised so that mean values were identical before and after adding $\gamma$). What $\sigma_{\gamma}$ does change is the variation in component interaction strengths, and $\rho$. Introducing variation in $\gamma$ increases the total variation in the system, making it more likely that $\textbf{M} = \gamma\textbf{A}$ is unstable. Variation of the off-diagonal elements in $\textbf{M}$ is described by the joint variation of two random variables (to simplify the notation, $\sigma^{2}_{M}$ and $\sigma^{2}_{A}$ refer to variances of off-diagonal elements only),

\begin{equation} \label{var_full}
\sigma^{2}_{M} = \sigma^{2}_{A}\sigma^{2}_{\gamma} + \sigma^{2}_{A}E[\gamma_{i}]^{2}+\sigma^{2}_{\gamma}E[A_{ij}]^{2}.
\end{equation}

Note that in my simulations $E[\gamma_{i}] = 1$ and $E[A_{ij}] = 0$, so equation \ref{var_full} can be simplified,

\begin{equation} \label{var_reduced}
\sigma^{2}_{M} = \sigma^{2}_{A}(1 + \sigma^{2}_{\gamma}).
\end{equation}
